---
title: "Final Clean Up Code Revised"
author: "COMM3501"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
# ============================================================
# 0. Setup
# ============================================================

# If any package is missing, install it once in the Console, for example:
# install.packages(c("readxl", "tidyverse", "janitor", "lubridate",
#                    "caret", "pROC", "e1071", "kknn", "rpart",
#                    "randomForest", "xgboost", "class", "knitr"))

library(readxl)
library(tidyverse)
library(janitor)
library(lubridate)
library(stringr)
library(forcats)
library(caret)
library(pROC)
library(e1071)
library(kknn)
library(rpart)
library(randomForest)
library(xgboost)
library(class)
library(knitr)
library(kernelshap)

SEED <- 123
TRAIN_PROPORTION <- 0.70
CLASSIFICATION_THRESHOLD <- 0.50

set.seed(SEED)

# Mac-safe absolute path. Change this if your file moves.
input_file <- "/Users/ch12yl/Downloads/2026T2/COMM3501/Assignment/A3 Group/A3_Dataset_2023.xlsx"

# Outputs will be saved beside this Rmd after knitting.
output_dir <- "model_outputs"
dir.create(output_dir, showWarnings = FALSE)
```

# Data Check

```{r data-check}
file.exists(input_file)

data_raw <- read_excel(input_file)

data_check <- tibble(
  Check = c(
    "Excel file exists",
    "Excel file opens in R on Mac",
    "Raw rows",
    "Raw columns"
  ),
  Result = c(
    ifelse(file.exists(input_file), "Pass", "Fail"),
    "Pass",
    nrow(data_raw),
    ncol(data_raw)
  )
)

kable(data_check)
```

# Clean Data

```{r clean-data, message=FALSE, warning=FALSE}
# 1. Clean column names and text
data <- data_raw %>%
  clean_names() %>%
  mutate(
    across(where(is.character), ~ str_squish(.)),
    across(where(is.character), ~ na_if(., "")),
    across(where(is.character), ~ na_if(., "[Blank]"))
  )

# 2. Check ID fields before removing them from modelling data
# IDs are kept until after the train/test split so duplicate records can be audited.
id_duplicate_check <- data %>%
  summarise(
    total_rows = n(),
    duplicate_recommendation_id = sum(duplicated(recommendation_id), na.rm = TRUE),
    duplicate_request_id = sum(duplicated(request_id), na.rm = TRUE),
    duplicate_life_id = sum(duplicated(life_id), na.rm = TRUE)
  ) %>%
  pivot_longer(
    everything(),
    names_to = "Check",
    values_to = "Value"
  )

data_clean <- data

# 3. Create target variable
# Positive class is "Yes", meaning the underwriter is NEOS Life.
data_clean <- data_clean %>%
  mutate(
    neos_flag = ifelse(underwriter == "NEOS Life", "Yes", "No"),
    neos_flag = factor(neos_flag, levels = c("Yes", "No"))
  )

# 4. Convert product Yes/No variables to binary
data_clean <- data_clean %>%
  mutate(
    life_bin = case_when(life == "Yes" ~ 1, life == "No" ~ 0, TRUE ~ NA_real_),
    tpd_bin = case_when(tpd == "Yes" ~ 1, tpd == "No" ~ 0, TRUE ~ NA_real_),
    trauma_bin = case_when(trauma == "Yes" ~ 1, trauma == "No" ~ 0, TRUE ~ NA_real_),
    ip_bin = case_when(ip == "Yes" ~ 1, ip == "No" ~ 0, TRUE ~ NA_real_),
    be_bin = case_when(be == "Yes" ~ 1, be == "No" ~ 0, TRUE ~ NA_real_),
    severity_bin = case_when(severity == "Yes" ~ 1, severity == "No" ~ 0, TRUE ~ NA_real_)
  )

# 5. Keep records with at least one core product
# The assignment focuses only on Life, TPD, Trauma and IP recommendations.
data_clean <- data_clean %>%
  mutate(
    product_count = rowSums(
      across(c(life_bin, tpd_bin, trauma_bin, ip_bin), ~ replace_na(.x, 0))
    )
  ) %>%
  filter(product_count > 0)

# 6. Clean numeric values
data_clean <- data_clean %>%
  mutate(
    annual_income = ifelse(annual_income < 0, NA_real_, annual_income),
    premium = ifelse(premium < 0, NA, premium),
    annualised_premium = ifelse(annualised_premium < 0, NA, annualised_premium),
    inside_super_premium = ifelse(inside_super_premium < 0, NA, inside_super_premium),
    outside_super_premium = ifelse(outside_super_premium < 0, NA, outside_super_premium)
  )

# 7. Create engineered variables
# Life, TPD and Trauma are lump-sum style covers.
# IP is periodic cover, so it is kept separately.
data_clean <- data_clean %>%
  mutate(
    has_alternative = ifelse(is.na(alternative), 0, 1),
    life_cover_specific = ifelse(life_bin == 1, replace_na(life_cover_amount, 0), 0),
    tpd_cover_specific = ifelse(tpd_bin == 1, replace_na(tpd_cover_amount, 0), 0),
    trauma_cover_specific = ifelse(trauma_bin == 1, replace_na(trauma_cover_amount, 0), 0),
    ip_cover_specific = ifelse(ip_bin == 1, replace_na(ip_cover_amount, 0), 0),
    lump_sum_cover = rowSums(
      across(
        c(life_cover_amount, tpd_cover_amount, trauma_cover_amount),
        ~ replace_na(.x, 0)
      )
    ),
    ip_periodic_cover = replace_na(ip_cover_amount, 0),
    log_annual_income = log1p(annual_income),
    log_premium = log1p(premium),
    log_annualised_premium = log1p(annualised_premium),
    log_life_cover = log1p(life_cover_specific),
    log_tpd_cover = log1p(tpd_cover_specific),
    log_trauma_cover = log1p(trauma_cover_specific),
    log_ip_cover = log1p(ip_cover_specific),
    log_lump_sum_cover = log1p(lump_sum_cover),
    log_ip_periodic_cover = log1p(ip_periodic_cover)
  )

# 8. Product bundle
data_clean <- data_clean %>%
  mutate(
    product_bundle = case_when(
      life_bin == 1 & tpd_bin == 0 & trauma_bin == 0 & ip_bin == 0 ~ "Life only",
      life_bin == 0 & tpd_bin == 1 & trauma_bin == 0 & ip_bin == 0 ~ "TPD only",
      life_bin == 0 & tpd_bin == 0 & trauma_bin == 1 & ip_bin == 0 ~ "Trauma only",
      life_bin == 0 & tpd_bin == 0 & trauma_bin == 0 & ip_bin == 1 ~ "IP only",
      life_bin == 1 & tpd_bin == 1 & trauma_bin == 0 & ip_bin == 0 ~ "Life + TPD",
      life_bin == 1 & tpd_bin == 1 & trauma_bin == 1 & ip_bin == 0 ~ "Life + TPD + Trauma",
      life_bin == 1 & tpd_bin == 1 & trauma_bin == 1 & ip_bin == 1 ~ "Life + TPD + Trauma + IP",
      product_count >= 2 ~ "Other bundle",
      TRUE ~ "Other"
    )
  )

# 9. Commission structure organisation
# Based on Australian life insurance 2020 legislation change.
# Creates 3 new columns:
# 1. commission_type
# 2. initial_commission_rate_gst
# 3. renewal_commission_rate_gst

commission_text <- data_clean$commission_structure %>%
  as.character() %>%
  str_squish() %>%
  str_to_lower() %>%
  replace_na("unknown")

# Extract C-rate, e.g. C100, C70, C85
c_rate <- suppressWarnings(
  as.numeric(str_match(commission_text, "c(\\d+(?:\\.\\d+)?)")[, 2])
)

# Extract direct pair, e.g. 66/22, 27.5/27.5, 40.7 / 16.5
pair_rate <- str_match(
  commission_text,
  "(\\d+(?:\\.\\d+)?)\\s*%?\\s*/\\s*(\\d+(?:\\.\\d+)?)\\s*%?"
)

pair_initial <- suppressWarnings(as.numeric(pair_rate[, 2]))
pair_renewal <- suppressWarnings(as.numeric(pair_rate[, 3]))

# Extract modifier pair, e.g. 100% Init / 100% Renew
init_modifier <- suppressWarnings(
  as.numeric(
    coalesce(
      str_match(commission_text, "(\\d+(?:\\.\\d+)?)\\s*%?\\s*(init|initial|year1|yr1)")[, 2],
      str_match(commission_text, "(init|initial|year1|yr1)\\s*(\\d+(?:\\.\\d+)?)\\s*%?")[, 3]
    )
  )
)

renew_modifier <- suppressWarnings(
  as.numeric(
    coalesce(
      str_match(commission_text, "(\\d+(?:\\.\\d+)?)\\s*%?\\s*(renew|renewal|year2|yr2)")[, 2],
      str_match(commission_text, "(renew|renewal|year2|yr2)\\s*(\\d+(?:\\.\\d+)?)\\s*%?")[, 3]
    )
  )
)

commission_type <- case_when(
  commission_text == "unknown" ~ "Unknown",
  str_detect(commission_text, "nil|nill|no commission|nil commission") ~ "No commission",
  str_detect(commission_text, "level|l2evel|0% level") ~ "Level",
  str_detect(commission_text, "upfront|higher initial|standard - upfront|initial \\(") ~ "Upfront / Higher initial",
  str_detect(commission_text, "hybrid|h4ybrid|h3ybrid|h2ybrid|hybrid60|hybrid70|hybrid80") ~ "Hybrid",
  TRUE ~ "Other specified"
)

# Base commission rates including GST
# 2020 LIF cap: 60% initial + GST = 66%; 20% renewal + GST = 22%
# Level commission is treated as 33% including GST.
base_initial <- case_when(
  commission_type == "Unknown" ~ NA_real_,
  commission_type == "No commission" ~ 0,
  commission_type == "Level" ~ 33,
  str_detect(commission_text, "2018|hybrid80|88\\s*/\\s*22") ~ 88,
  str_detect(commission_text, "2019|hybrid70|77\\s*/\\s*22") ~ 77,
  commission_type %in% c("Hybrid", "Upfront / Higher initial") ~ 66,
  TRUE ~ NA_real_
)

base_renewal <- case_when(
  commission_type == "Unknown" ~ NA_real_,
  commission_type == "No commission" ~ 0,
  commission_type == "Level" ~ 33,
  commission_type %in% c("Hybrid", "Upfront / Higher initial") ~ 22,
  TRUE ~ NA_real_
)

# Decide whether pair values are actual rates or modifiers.
# Example: 66/22 is actual.
# Example: 100 Init / 100 Renew is a modifier of the base rate.
has_modifier_words <- str_detect(
  commission_text,
  "init|initial|renew|renewal|year1|year2"
)

use_pair_as_modifier <- has_modifier_words &
  !str_detect(commission_text, "yr1\\s*66|year1\\s*66|yr2\\s*22|year2\\s*22") &
  !is.na(pair_initial) &
  !is.na(pair_renewal) &
  pair_initial <= 100 &
  pair_renewal <= 100

initial_rate <- case_when(
  commission_type == "Unknown" ~ NA_real_,
  commission_type == "No commission" ~ 0,
  !is.na(c_rate) ~ base_initial * c_rate / 100,
  !is.na(init_modifier) ~ base_initial * init_modifier / 100,
  use_pair_as_modifier ~ base_initial * pair_initial / 100,
  !is.na(pair_initial) ~ pair_initial,
  commission_type %in% c("Hybrid", "Upfront / Higher initial", "Level") ~ base_initial,
  TRUE ~ NA_real_
)

renewal_rate <- case_when(
  commission_type == "Unknown" ~ NA_real_,
  commission_type == "No commission" ~ 0,
  !is.na(c_rate) ~ base_renewal * c_rate / 100,
  !is.na(renew_modifier) ~ base_renewal * renew_modifier / 100,
  use_pair_as_modifier ~ base_renewal * pair_renewal / 100,
  !is.na(pair_renewal) ~ pair_renewal,
  commission_type %in% c("Hybrid", "Upfront / Higher initial", "Level") ~ base_renewal,
  TRUE ~ NA_real_
)

data_clean <- data_clean %>%
  mutate(
    commission_type = factor(commission_type),
    initial_commission_rate_gst = case_when(
      commission_text == "unknown" ~ "Unknown",
      is.na(initial_rate) ~ "Not stated in label",
      TRUE ~ paste0(round(initial_rate, 2), "%")
    ),
    renewal_commission_rate_gst = case_when(
      commission_text == "unknown" ~ "Unknown",
      is.na(renewal_rate) ~ "Not stated in label",
      TRUE ~ paste0(round(renewal_rate, 2), "%")
    ),
    initial_commission_rate_gst = factor(initial_commission_rate_gst),
    renewal_commission_rate_gst = factor(renewal_commission_rate_gst)
  )

# 10. Date variables
fix_excel_date <- function(x) {
  if (inherits(x, "Date")) return(x)
  if (inherits(x, "POSIXct")) return(as.Date(x))
  as.Date(as.numeric(x), origin = "1899-12-30")
}

data_clean <- data_clean %>%
  mutate(
    date = fix_excel_date(date),
    year = year(date),
    month = month(date, label = TRUE),
    quarter = quarter(date),
    weekday = wday(date, label = TRUE),
    age_band = case_when(
      age_next < 30 ~ "Under 30",
      age_next < 40 ~ "30-39",
      age_next < 50 ~ "40-49",
      age_next < 60 ~ "50-59",
      age_next >= 60 ~ "60+",
      TRUE ~ NA_character_
    )
  )

# 11. Collapse high-cardinality categorical variables
data_clean <- data_clean %>%
  mutate(
    occupation_group = fct_lump_n(as.factor(occupation), n = 30, other_level = "Other"),
    package_group = fct_lump_n(as.factor(package), n = 30, other_level = "Other"),
    adviser_group = fct_lump_n(as.factor(adviser_id), n = 50, other_level = "Other"),
    occupation_text = str_to_lower(as.character(occupation)),
    occupation_category = case_when(
      str_detect(occupation_text, "home duties|retired|student|unemployed") ~ "Home / retired / student",
      str_detect(occupation_text, "doctor|medical|nurse|dentist|physio|psychologist|pharmacist|veterinary|surgeon|therapist|paramedic") ~ "Healthcare professional",
      str_detect(occupation_text, "teacher|education|lecturer|academic|child care|teachers aide") ~ "Education / childcare",
      str_detect(occupation_text, "lawyer|solicitor|legal|accountant|architect|engineer|computer|programmer|analyst|consultant|actuary|scientist") ~ "Professional / technical",
      str_detect(occupation_text, "manager|management|chief executive|project manager|business development") ~ "Management",
      str_detect(occupation_text, "clerical|administration|clerk|receptionist|bookkeeper|office|bank") ~ "Clerical / administration",
      str_detect(occupation_text, "sales|real estate|retail|marketing") ~ "Sales / marketing",
      str_detect(occupation_text, "electrician|plumber|carpenter|builder|mechanic|fitter|cabinet|chef|hairdresser|trade|construction|foreman") ~ "Trade / skilled manual",
      str_detect(occupation_text, "driver|truck|mining|farming|police|fire|security|plant operator|manual|blue collar|heavy") ~ "Manual / field / higher risk",
      str_detect(occupation_text, "1a|1b|white collar|1p|1l|1m") ~ "Professional / technical",
      str_detect(occupation_text, "2a|2b|2c") ~ "White collar / clerical",
      str_detect(occupation_text, "3a|3b|3m|\\b4\\b|\\b5\\b") ~ "Manual / field / higher risk",
      TRUE ~ "Other"
    ),
    occupation_category = fct_lump_min(
      as.factor(occupation_category),
      min = 50,
      other_level = "Other"
    )
  )

# 12. Convert categorical fields to factors
data_clean <- data_clean %>%
  mutate(
    across(
      c(
        neos_flag, underwriter, package, package_group,
        super, rollover_tax_rebate, gender, smoker_status,
        home_state, occupation, occupation_group, occupation_category, self_employed,
        premium_frequency, product_bundle, commission_type,
        initial_commission_rate_gst, renewal_commission_rate_gst,
        age_band, month, weekday, adviser_group
      ),
      as.factor
    )
  )
```

# Missing Values

```{r missing-values}
id_duplicate_check %>%
  kable(caption = "ID duplicate audit before ID removal")

product_scope_summary <- data_clean %>%
  summarise(
    Life = sum(life_bin == 1, na.rm = TRUE),
    TPD = sum(tpd_bin == 1, na.rm = TRUE),
    Trauma = sum(trauma_bin == 1, na.rm = TRUE),
    IP = sum(ip_bin == 1, na.rm = TRUE),
    Multi_Product = sum(product_count >= 2, na.rm = TRUE)
  ) %>%
  pivot_longer(
    everything(),
    names_to = "Product_Group",
    values_to = "Recommendation_Count"
  ) %>%
  mutate(
    Percent = round(Recommendation_Count / nrow(data_clean) * 100, 2)
  )

product_scope_summary %>%
  kable(caption = "Life, TPD, Trauma and IP product scope")

missing_table <- data_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Missing_Count"
  ) %>%
  mutate(
    Missing_Percent = round(Missing_Count / nrow(data_clean) * 100, 4)
  ) %>%
  arrange(desc(Missing_Count), Variable)

kable(missing_table)
```

# Final Modelling Data

```{r model-data}
id_columns <- c(
  "row_id",
  "recommendation_id",
  "request_id",
  "life_id",
  "external_ref"
)

# Shared modelling population for all supervised models.
# IDs stay here until after train/test splitting, then they are removed.
model_data_with_ids <- data_clean %>%
  mutate(
    row_id = row_number(),
    has_alternative = factor(has_alternative, levels = c(0, 1))
  ) %>%
  dplyr::select(
    row_id,
    recommendation_id,
    request_id,
    life_id,
    external_ref,
    neos_flag,

    # Customer characteristics
    age_next,
    gender,
    smoker_status,
    home_state,
    occupation_group,
    occupation_category,
    self_employed,
    log_annual_income,

    # Product characteristics
    super,
    rollover_tax_rebate,
    life_bin,
    tpd_bin,
    trauma_bin,
    ip_bin,
    product_count,
    product_bundle,
    log_annualised_premium,
    log_lump_sum_cover,
    log_ip_periodic_cover,
    premium_frequency,

    # Adviser and recommendation characteristics.
    # Commission is excluded from final predictors after split due to leakage risk.
    has_alternative
  ) %>%
  filter(if_all(-all_of(id_columns), ~ !is.na(.x)))

naive_data_final <- model_data_with_ids %>%
  dplyr::select(
    all_of(id_columns),
    neos_flag,
    age_next, gender, smoker_status, home_state,
    occupation_group, self_employed, log_annual_income,
    super, rollover_tax_rebate,
    life_bin, tpd_bin, trauma_bin, ip_bin, product_count, product_bundle,
    log_annualised_premium, log_lump_sum_cover,
    log_ip_periodic_cover,
    premium_frequency,
    has_alternative
  ) %>%
  mutate(neos_flag = factor(neos_flag, levels = c("Yes", "No")))

knn_data_final <- model_data_with_ids %>%
  dplyr::select(
    all_of(id_columns),
    neos_flag,
    age_next, gender, smoker_status, home_state,
    occupation_category, self_employed, log_annual_income,
    super, rollover_tax_rebate,
    life_bin, tpd_bin, trauma_bin, ip_bin, product_count,
    log_annualised_premium, log_lump_sum_cover,
    log_ip_periodic_cover,
    premium_frequency,
    has_alternative
  ) %>%
  mutate(neos_flag = factor(neos_flag, levels = c("Yes", "No")))

tree_data_final <- model_data_with_ids %>%
  dplyr::select(
    all_of(id_columns),
    neos_flag,
    age_next, gender, smoker_status, home_state,
    occupation_group, self_employed, log_annual_income,
    super, rollover_tax_rebate,
    life_bin, tpd_bin, trauma_bin, ip_bin, product_count, product_bundle,
    log_annualised_premium, log_lump_sum_cover,
    log_ip_periodic_cover,
    premium_frequency,
    has_alternative
  ) %>%
  mutate(neos_flag = factor(neos_flag, levels = c("Yes", "No")))

common_model_ids <- Reduce(
  intersect,
  list(
    naive_data_final$row_id,
    knn_data_final$row_id,
    tree_data_final$row_id
  )
)

model_data <- tree_data_final %>%
  filter(row_id %in% common_model_ids) %>%
  arrange(row_id)

naive_data_final <- naive_data_final %>%
  filter(row_id %in% common_model_ids) %>%
  arrange(row_id)

knn_data_final <- knn_data_final %>%
  filter(row_id %in% common_model_ids) %>%
  arrange(row_id)

tree_data_final <- tree_data_final %>%
  filter(row_id %in% common_model_ids) %>%
  arrange(row_id)

product_model_data <- list(
  Life = model_data %>% filter(life_bin == 1),
  TPD = model_data %>% filter(tpd_bin == 1),
  Trauma = model_data %>% filter(trauma_bin == 1),
  IP = model_data %>% filter(ip_bin == 1)
)

product_model_summary <- imap_dfr(
  product_model_data,
  ~ tibble(
    Product = .y,
    Records = nrow(.x),
    NEOS_Yes_Rate = mean(.x$neos_flag == "Yes")
  )
)

# Check the number of observations and class balance.
model_population_summary <- bind_rows(
  naive_data_final %>% count(neos_flag) %>% mutate(Model_Data = "Naive / Complement Naive"),
  knn_data_final %>% count(neos_flag) %>% mutate(Model_Data = "KNN"),
  tree_data_final %>% count(neos_flag) %>% mutate(Model_Data = "Logistic / Tree / Ensemble")
) %>%
  group_by(Model_Data) %>%
  mutate(percent = n / sum(n))

kable(model_population_summary, caption = "Final modelling data class balance")

kable(product_model_summary, caption = "Product-specific modelling subsets")

# Save one reusable handoff file for downstream modelling documents. This lets
# later RMDs load the prepared datasets without repeating data cleaning.
prepared_modelling_objects <- list(
  model_data = model_data,
  naive_data_final = naive_data_final,
  knn_data_final = knn_data_final,
  tree_data_final = tree_data_final,
  product_model_data = product_model_data,
  product_model_summary = product_model_summary,
  model_population_summary = model_population_summary,
  id_columns = id_columns
)

saveRDS(
  prepared_modelling_objects,
  file.path(output_dir, "prepared_modelling_objects.rds")
)
```

# Helper Functions

```{r modelling-helpers}
# Convert predicted probabilities into Yes/No classifications.
probability_to_class <- function(
    probability,
    threshold = CLASSIFICATION_THRESHOLD
) {
  factor(
    if_else(probability >= threshold, "Yes", "No"),
    levels = c("Yes", "No")
  )
}

# Evaluate model performance.
make_confusion_matrix <- function(
    predicted,
    actual
) {
  predicted <- factor(
    as.character(predicted),
    levels = c("Yes", "No")
  )

  actual <- factor(
    as.character(actual),
    levels = c("Yes", "No")
  )

  confusionMatrix(
    data = predicted,
    reference = actual,
    positive = "Yes"
  )
}

evaluate_model <- function(
    model_name,
    actual,
    predicted,
    probability
) {

  actual <- factor(as.character(actual), levels = c("Yes", "No"))
  predicted <- factor(as.character(predicted), levels = c("Yes", "No"))

  confusion <- make_confusion_matrix(
    predicted = predicted,
    actual = actual
  )

  roc_result <- roc(
    response = actual,
    predictor = probability,
    levels = c("No", "Yes"),
    direction = "<",
    quiet = TRUE
  )

  tibble(
    Model = model_name,
    Accuracy = as.numeric(
      confusion$overall["Accuracy"]
    ),
    Sensitivity = as.numeric(
      confusion$byClass["Sensitivity"]
    ),
    Specificity = as.numeric(
      confusion$byClass["Specificity"]
    ),
    Precision = as.numeric(
      confusion$byClass["Pos Pred Value"]
    ),
    F1_Score = as.numeric(
      confusion$byClass["F1"]
    ),
    Balanced_Accuracy = as.numeric(
      confusion$byClass["Balanced Accuracy"]
    ),
    AUC = as.numeric(
      auc(roc_result)
    )
  )
}

make_split_ids <- function(data, p = TRAIN_PROPORTION) {
  train_index <- createDataPartition(data$neos_flag, p = p, list = FALSE)
  list(
    train_ids = data$row_id[train_index],
    test_ids = data$row_id[-train_index]
  )
}

sample_stratified <- function(data, cap) {
  if (is.infinite(cap) || nrow(data) <= cap) return(data)
  sample_index <- createDataPartition(
    data$neos_flag,
    p = cap / nrow(data),
    list = FALSE
  )
  data[sample_index, ]
}

performance_metrics <- function(model_name, actual, predicted_class, predicted_prob) {
  cm <- confusionMatrix(predicted_class, actual, positive = "Yes")
  auc_value <- auc(
    response = actual,
    predictor = predicted_prob,
    levels = c("No", "Yes"),
    quiet = TRUE
  )

  tibble(
    Model = model_name,
    AUC = as.numeric(auc_value),
    Accuracy = unname(cm$overall["Accuracy"]),
    Kappa = unname(cm$overall["Kappa"]),
    Sensitivity_Recall = unname(cm$byClass["Sensitivity"]),
    Specificity = unname(cm$byClass["Specificity"]),
    Precision = unname(cm$byClass["Precision"]),
    F1 = unname(cm$byClass["F1"]),
    Balanced_Accuracy = unname(cm$byClass["Balanced Accuracy"])
  )
}

confusion_table <- function(model_name, actual, predicted_class) {
  cm <- confusionMatrix(predicted_class, actual, positive = "Yes")

  as.data.frame.matrix(cm$table) %>%
    rownames_to_column("Predicted") %>%
    mutate(Model = model_name, .before = 1)
}

lift_table <- function(model_name, actual, predicted_prob) {
  base_rate <- mean(actual == "Yes")

  tibble(
    Actual = actual,
    Probability_Yes = predicted_prob
  ) %>%
    arrange(desc(Probability_Yes)) %>%
    mutate(
      Decile = ntile(-Probability_Yes, 10),
      Is_Yes = as.integer(Actual == "Yes")
    ) %>%
    group_by(Decile) %>%
    summarise(
      Model = model_name,
      Records = n(),
      Actual_Yes = sum(Is_Yes),
      Mean_Probability_Yes = mean(Probability_Yes),
      Response_Rate = mean(Is_Yes),
      Decile_Lift = Response_Rate / base_rate,
      .groups = "drop"
    ) %>%
    arrange(Decile) %>%
    mutate(
      Cumulative_Records = cumsum(Records),
      Cumulative_Yes = cumsum(Actual_Yes),
      Cumulative_Response_Rate = Cumulative_Yes / Cumulative_Records,
      Cumulative_Lift = Cumulative_Response_Rate / base_rate
    )
}

best_roc_threshold <- function(actual, predicted_prob) {
  roc_object <- roc(
    response = actual,
    predictor = predicted_prob,
    levels = c("No", "Yes"),
    quiet = TRUE
  )

  as.numeric(coords(
    roc_object,
    x = "best",
    best.method = "youden",
    ret = "threshold"
  ))
}

fit_complement_nb <- function(train_data, target = "neos_flag", alpha = 1) {
  y <- train_data[[target]]
  x_raw <- train_data %>% dplyr::select(-all_of(target))

  numeric_cols <- names(x_raw)[map_lgl(x_raw, is.numeric)]
  numeric_stats <- map(
    numeric_cols,
    ~ list(
      min = min(x_raw[[.x]], na.rm = TRUE),
      max = max(x_raw[[.x]], na.rm = TRUE)
    )
  )
  names(numeric_stats) <- numeric_cols

  x_scaled <- x_raw
  for (col_name in numeric_cols) {
    col_min <- numeric_stats[[col_name]]$min
    col_max <- numeric_stats[[col_name]]$max
    if (is.finite(col_min) && is.finite(col_max) && col_max > col_min) {
      x_scaled[[col_name]] <- (x_scaled[[col_name]] - col_min) / (col_max - col_min)
    } else {
      x_scaled[[col_name]] <- 0
    }
  }

  dummy_model <- dummyVars(~ ., data = x_scaled, fullRank = FALSE)
  x_matrix <- as.matrix(predict(dummy_model, newdata = x_scaled))
  x_matrix[is.na(x_matrix)] <- 0
  x_matrix[x_matrix < 0] <- 0

  class_levels <- levels(y)
  class_prior <- table(y)[class_levels] / length(y)

  log_theta <- map_dfr(class_levels, function(class_name) {
    complement_x <- x_matrix[y != class_name, , drop = FALSE]
    feature_count <- colSums(complement_x) + alpha
    feature_prob <- feature_count / sum(feature_count)
    tibble(Class = class_name, Feature = names(feature_prob), Log_Theta = log(feature_prob))
  })

  list(
    dummy_model = dummy_model,
    numeric_stats = numeric_stats,
    feature_names = colnames(x_matrix),
    log_theta = log_theta,
    class_levels = class_levels,
    class_prior = class_prior,
    target = target
  )
}

predict_complement_nb <- function(model, newdata) {
  x_raw <- newdata %>% dplyr::select(-all_of(model$target))
  x_scaled <- x_raw

  for (col_name in names(model$numeric_stats)) {
    col_min <- model$numeric_stats[[col_name]]$min
    col_max <- model$numeric_stats[[col_name]]$max
    if (is.finite(col_min) && is.finite(col_max) && col_max > col_min) {
      x_scaled[[col_name]] <- (x_scaled[[col_name]] - col_min) / (col_max - col_min)
      x_scaled[[col_name]] <- pmin(pmax(x_scaled[[col_name]], 0), 1)
    } else {
      x_scaled[[col_name]] <- 0
    }
  }

  x_matrix <- as.matrix(predict(model$dummy_model, newdata = x_scaled))
  x_matrix[is.na(x_matrix)] <- 0
  x_matrix[x_matrix < 0] <- 0

  missing_features <- setdiff(model$feature_names, colnames(x_matrix))
  if (length(missing_features) > 0) {
    x_matrix <- cbind(
      x_matrix,
      matrix(0, nrow = nrow(x_matrix), ncol = length(missing_features),
             dimnames = list(NULL, missing_features))
    )
  }
  x_matrix <- x_matrix[, model$feature_names, drop = FALSE]

  log_theta_matrix <- model$log_theta %>%
    pivot_wider(names_from = Feature, values_from = Log_Theta) %>%
    arrange(match(Class, model$class_levels))

  theta_matrix <- as.matrix(log_theta_matrix %>% dplyr::select(-Class))
  theta_matrix <- theta_matrix[, model$feature_names, drop = FALSE]

  scores <- -x_matrix %*% t(theta_matrix)
  scores <- sweep(scores, 2, log(as.numeric(model$class_prior[model$class_levels])), "+")
  colnames(scores) <- model$class_levels

  scores <- sweep(scores, 1, apply(scores, 1, max), "-")
  prob <- exp(scores)
  prob <- prob / rowSums(prob)
  as.data.frame(prob)
}

knn_summary <- function(data, lev = NULL, model = NULL) {
  roc_stats <- twoClassSummary(data, lev = lev, model = model)
  accuracy <- mean(data$pred == data$obs)
  c(roc_stats, Accuracy = accuracy)
}
```

# Train-Test Split

```{r split-data}
# One shared split for all models.
# IDs are used only to align rows, then removed from model predictors.
set.seed(SEED)

train_index <- createDataPartition(
  y = model_data$neos_flag,
  p = TRAIN_PROPORTION,
  list = FALSE
)

shared_train_ids <- model_data$row_id[train_index]
shared_test_ids <- model_data$row_id[-train_index]

naive_train <- naive_data_final %>%
  filter(row_id %in% shared_train_ids) %>%
  dplyr::select(-all_of(id_columns))

naive_test <- naive_data_final %>%
  filter(row_id %in% shared_test_ids) %>%
  dplyr::select(-all_of(id_columns))

knn_train_base <- knn_data_final %>%
  filter(row_id %in% shared_train_ids)

knn_test_base <- knn_data_final %>%
  filter(row_id %in% shared_test_ids)

train_data <- tree_data_final %>%
  filter(row_id %in% shared_train_ids) %>%
  dplyr::select(-all_of(id_columns))

test_data <- tree_data_final %>%
  filter(row_id %in% shared_test_ids) %>%
  dplyr::select(-all_of(id_columns))

split_summary <- tibble(
  Dataset = c("Training", "Testing"),
  Rows = c(nrow(train_data), nrow(test_data)),
  NEOS_Yes_Rate = c(
    mean(train_data$neos_flag == "Yes"),
    mean(test_data$neos_flag == "Yes")
  )
)

kable(split_summary, caption = "Shared train-test split summary")
```

# Naive Bayes

```{r naive-bayes-model, message=FALSE, warning=FALSE}
naive_model <- naiveBayes(
  neos_flag ~ .,
  data = naive_train,
  laplace = 1
)

naive_train_probability <- predict(
  naive_model,
  newdata = naive_train,
  type = "raw"
)[, "Yes"]

naive_threshold <- best_roc_threshold(naive_train$neos_flag, naive_train_probability)

naive_probability <- predict(
  naive_model,
  newdata = naive_test,
  type = "raw"
)[, "Yes"]

naive_prediction <- probability_to_class(
  naive_probability,
  naive_threshold
)
```

# Complement Naive Bayes

```{r complement-naive-bayes-model, message=FALSE, warning=FALSE}
complement_naive_model <- fit_complement_nb(
  train_data = naive_train,
  target = "neos_flag",
  alpha = 1
)

complement_naive_train_probability <- predict_complement_nb(
  complement_naive_model,
  naive_train
)[, "Yes"]

complement_naive_threshold <- best_roc_threshold(
  naive_train$neos_flag,
  complement_naive_train_probability
)

complement_naive_probability <- predict_complement_nb(
  complement_naive_model,
  naive_test
)[, "Yes"]

complement_naive_prediction <- probability_to_class(
  complement_naive_probability,
  complement_naive_threshold
)
```

# KNN

```{r knn-model, message=FALSE, warning=FALSE}
knn_train_cap <- 1000
knn_test_cap <- Inf

knn_train_sample <- knn_train_base %>%
  sample_stratified(knn_train_cap)

knn_test_sample <- knn_test_base %>%
  sample_stratified(knn_test_cap)

knn_train_ids <- knn_train_sample$row_id
knn_test_ids <- knn_test_sample$row_id

knn_train <- knn_train_sample %>%
  dplyr::select(-all_of(id_columns))

knn_test <- knn_test_sample %>%
  dplyr::select(-all_of(id_columns))

knn_dummy_model <- dummyVars(
  ~ .,
  data = knn_train %>% dplyr::select(-neos_flag),
  fullRank = TRUE
)

knn_x_train <- as.data.frame(
  predict(knn_dummy_model, newdata = knn_train %>% dplyr::select(-neos_flag))
)

knn_x_test <- as.data.frame(
  predict(knn_dummy_model, newdata = knn_test %>% dplyr::select(-neos_flag))
)

knn_nzv_cols <- nearZeroVar(knn_x_train)
if (length(knn_nzv_cols) > 0) {
  knn_x_train <- knn_x_train[, -knn_nzv_cols, drop = FALSE]
  knn_x_test <- knn_x_test[, names(knn_x_train), drop = FALSE]
}

knn_cor_matrix <- cor(knn_x_train)
knn_high_corr <- findCorrelation(knn_cor_matrix, cutoff = 0.90, names = TRUE)

if (length(knn_high_corr) > 0) {
  knn_x_train <- knn_x_train[, !names(knn_x_train) %in% knn_high_corr, drop = FALSE]
  knn_x_test <- knn_x_test[, names(knn_x_train), drop = FALSE]
}

knn_scale_model <- preProcess(knn_x_train, method = c("center", "scale"))
knn_x_train <- predict(knn_scale_model, knn_x_train)
knn_x_test <- predict(knn_scale_model, knn_x_test)

knn_train_processed <- cbind(neos_flag = knn_train$neos_flag, knn_x_train)
knn_test_processed <- cbind(neos_flag = knn_test$neos_flag, knn_x_test)

knn_control_base <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = knn_summary
)

knn_control_down <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = knn_summary,
  sampling = "down"
)

knn_grid <- expand.grid(
  kmax = c(5, 11, 21, 31, 51),
  distance = 2,
  kernel = c("rectangular", "triangular")
)

knn_model_base <- train(
  neos_flag ~ .,
  data = knn_train_processed,
  method = "kknn",
  metric = "ROC",
  trControl = knn_control_base,
  tuneGrid = knn_grid
)

knn_model_down <- train(
  neos_flag ~ .,
  data = knn_train_processed,
  method = "kknn",
  metric = "ROC",
  trControl = knn_control_down,
  tuneGrid = knn_grid
)

knn_base_train_probability <- predict(
  knn_model_base,
  newdata = knn_train_processed,
  type = "prob"
)[, "Yes"]

knn_down_train_probability <- predict(
  knn_model_down,
  newdata = knn_train_processed,
  type = "prob"
)[, "Yes"]

knn_base_threshold <- best_roc_threshold(knn_train$neos_flag, knn_base_train_probability)
knn_down_threshold <- best_roc_threshold(knn_train$neos_flag, knn_down_train_probability)

knn_base_probability <- predict(
  knn_model_base,
  newdata = knn_test_processed,
  type = "prob"
)[, "Yes"]

knn_down_probability <- predict(
  knn_model_down,
  newdata = knn_test_processed,
  type = "prob"
)[, "Yes"]

knn_base_prediction <- probability_to_class(
  knn_base_probability,
  knn_base_threshold
)

knn_down_prediction <- probability_to_class(
  knn_down_probability,
  knn_down_threshold
)

knn_model_base$bestTune
knn_model_down$bestTune
```

# Logistic Regression

```{r logistic-regression-model, message=FALSE, warning=FALSE}
# A separate numeric response ensures that the model predicts
# the probability of "Yes" rather than the probability of "No".
logistic_train_data <- train_data %>%
  mutate(
    neos_numeric = if_else(neos_flag == "Yes", 1, 0)
  ) %>%
  dplyr::select(-neos_flag)

logistic_test_x <- test_data %>%
  dplyr::select(-neos_flag)

logistic_regression_model <- glm(
  neos_numeric ~ .,
  data = logistic_train_data,
  family = binomial(link = "logit")
)

logistic_regression_probability <- predict(
  logistic_regression_model,
  newdata = logistic_test_x,
  type = "response"
)

logistic_regression_prediction <- probability_to_class(
  logistic_regression_probability
)
```

# Decision Tree

```{r decision-tree-model, message=FALSE, warning=FALSE}
set.seed(SEED)

tree_tune_control <- trainControl(
  method = "cv",
  number = 3,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

decision_tree_model <- train(
  neos_flag ~ .,
  data = train_data,
  method = "rpart",
  metric = "ROC",
  trControl = tree_tune_control,
  tuneGrid = expand.grid(
    cp = c(0.001, 0.005, 0.01, 0.02)
  ),
  parms = list(
    prior = c(Yes = 0.50, No = 0.50),
    split = "gini"
  ),
  control = rpart.control(minsplit = 20, minbucket = 7, maxdepth = 10)
)

decision_tree_probability <- predict(
  decision_tree_model,
  newdata = test_data,
  type = "prob"
)[, "Yes"]

decision_tree_prediction <- probability_to_class(
  decision_tree_probability
)
```

# Random Forest

```{r random-forest-model, message=FALSE, warning=FALSE}
set.seed(SEED)

# Random Forest uses its own resampling control so this chunk does not depend
# on the Decision Tree control object. Saved out-of-fold probabilities are
# required for the later Max-F1 threshold analysis.
rf_control <- trainControl(
  method = "cv",
  number = 3,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

rf_mtry_grid <- expand.grid(
  mtry = sort(unique(pmin(
    ncol(train_data) - 1,
    pmax(
      1,
      c(
        2,
        4,
        6,
        8,
        floor(sqrt(ncol(train_data) - 1))
      )
    )
  )))
)

random_forest_model <- train(
  neos_flag ~ .,
  data = train_data,
  method = "rf",
  metric = "ROC",
  trControl = rf_control,
  tuneGrid = rf_mtry_grid,
  ntree = 100,
  importance = TRUE
)

# Report the cross-validated mtry selected by ROC-AUC.
random_forest_model$bestTune

# Prevent an older direct randomForest() object remaining in the R session from
# being saved under the new caret-model filename.
if (!inherits(random_forest_model, "train")) {
  stop(
    "random_forest_model is an older model in the current R session. Restart R and run the complete Random Forest chunk, including caret::train()."
  )
}

rf_model_predictors <- attr(
  random_forest_model$terms,
  "term.labels"
)

rf_missing_predictors <- setdiff(
  rf_model_predictors,
  names(test_data)
)

if (length(rf_missing_predictors) > 0) {
  stop(
    "The fitted Random Forest expects predictors not found in the current test data: ",
    paste(rf_missing_predictors, collapse = ", "),
    ". Restart R and run the data and Random Forest chunks again."
  )
}

# Save the current 100-tree caret model immediately after fitting. A distinct
# filename prevents an older direct randomForest() object from being reused.
saveRDS(
  random_forest_model,
  file.path(output_dir, "random_forest_caret_100_trees.rds")
)

random_forest_probability <- predict(
  random_forest_model,
  newdata = test_data,
  type = "prob"
)[, "Yes"]

random_forest_prediction <- probability_to_class(
  random_forest_probability
)
```

# XGBoost Data Preparation

```{r xgboost-data-preparation, message=FALSE, warning=FALSE}
# Commission variables are excluded from XGBoost so this model focuses on
# customer, product and recommendation profile signals only.
xgb_train_data <- train_data %>%
  dplyr::select(
    -matches("commission|initial_rate|renewal_rate")
  )

xgb_test_data <- test_data %>%
  dplyr::select(
    -matches("commission|initial_rate|renewal_rate")
  )

# XGBoost requires numeric predictors, so categorical variables
# are converted into dummy variables.
xgb_dummy_model <- dummyVars(
  neos_flag ~ .,
  data = xgb_train_data,
  fullRank = FALSE
)

xgb_train_x <- predict(
  xgb_dummy_model,
  newdata = xgb_train_data
) %>%
  as.matrix()

xgb_test_x <- predict(
  xgb_dummy_model,
  newdata = xgb_test_data
) %>%
  as.matrix()

xgb_train_y <- if_else(
  xgb_train_data$neos_flag == "Yes",
  1,
  0
)
```

# XGBoost

```{r xgboost-model, message=FALSE, warning=FALSE}
xgb_tune_grid <- expand.grid(
  nrounds = c(50, 100),
  max_depth = 3,
  eta = 0.10,
  gamma = 0,
  colsample_bytree = 0.80,
  min_child_weight = 1,
  subsample = 0.80
)

xgb_train_matrix <- xgb.DMatrix(
  data = xgb_train_x,
  label = xgb_train_y
)

xgb_test_matrix <- xgb.DMatrix(
  data = xgb_test_x
)

xgb_cv_results <- pmap_dfr(
  xgb_tune_grid,
  function(
    nrounds,
    max_depth,
    eta,
    gamma,
    colsample_bytree,
    min_child_weight,
    subsample
  ) {
    set.seed(SEED)

    cv_model <- xgb.cv(
      data = xgb_train_matrix,
      objective = "binary:logistic",
      eval_metric = "auc",
      nrounds = nrounds,
      nfold = 3,
      max_depth = max_depth,
      eta = eta,
      gamma = gamma,
      colsample_bytree = colsample_bytree,
      min_child_weight = min_child_weight,
      subsample = subsample,
      stratified = TRUE,
      verbose = 0,
      early_stopping_rounds = 10
    )

    cv_log <- cv_model$evaluation_log
    auc_column <- names(cv_log)[str_detect(names(cv_log), "test_auc_mean")]

    best_iteration <- if (!is.null(cv_model$best_iteration)) {
      cv_model$best_iteration
    } else {
      cv_log$iter[which.max(cv_log[[auc_column]])]
    }

    tibble(
      nrounds = nrounds,
      max_depth = max_depth,
      eta = eta,
      gamma = gamma,
      colsample_bytree = colsample_bytree,
      min_child_weight = min_child_weight,
      subsample = subsample,
      best_iteration = best_iteration,
      CV_AUC = max(cv_log[[auc_column]], na.rm = TRUE)
    )
  }
) %>%
  arrange(desc(CV_AUC))

xgb_best_params <- xgb_cv_results %>%
  slice_head(n = 1)

xgb_best_nrounds <- xgb_best_params$best_iteration

if (length(xgb_best_nrounds) == 0 || is.na(xgb_best_nrounds) || xgb_best_nrounds < 1) {
  xgb_best_nrounds <- xgb_best_params$nrounds
}

xgboost_model <- xgb.train(
  data = xgb_train_matrix,
  objective = "binary:logistic",
  eval_metric = "auc",
  nrounds = xgb_best_nrounds,
  max_depth = xgb_best_params$max_depth,
  eta = xgb_best_params$eta,
  gamma = xgb_best_params$gamma,
  colsample_bytree = xgb_best_params$colsample_bytree,
  min_child_weight = xgb_best_params$min_child_weight,
  subsample = xgb_best_params$subsample,
  verbose = 0
)

xgboost_probability <- predict(
  xgboost_model,
  newdata = xgb_test_matrix
)

xgboost_prediction <- probability_to_class(
  xgboost_probability
)

kable(xgb_cv_results, digits = 4, caption = "XGBoost cross-validated AUC tuning results")
```

# Model Comparison

```{r model-comparison}
evaluation_actual <- factor(
  as.character(test_data$neos_flag),
  levels = c("Yes", "No")
)

evaluation_dataset_check <- tibble(
  Evaluation_Set = c(
    "Shared test set",
    "Naive / Complement Naive test set",
    "KNN test set",
    "Logistic / Tree / Ensemble test set"
  ),
  Rows = c(
    length(evaluation_actual),
    nrow(naive_test),
    nrow(knn_test),
    nrow(test_data)
  ),
  NEOS_Yes_Rate = c(
    mean(evaluation_actual == "Yes"),
    mean(naive_test$neos_flag == "Yes"),
    mean(knn_test$neos_flag == "Yes"),
    mean(test_data$neos_flag == "Yes")
  )
)

kable(
  evaluation_dataset_check,
  digits = 4,
  caption = "Evaluation dataset consistency check"
)

model_comparison <- bind_rows(
  evaluate_model(
    model_name = "Naive Bayes",
    actual = evaluation_actual,
    predicted = naive_prediction,
    probability = naive_probability
  ),

  evaluate_model(
    model_name = "Complement Naive Bayes",
    actual = evaluation_actual,
    predicted = complement_naive_prediction,
    probability = complement_naive_probability
  ),

  evaluate_model(
    model_name = "KNN baseline",
    actual = evaluation_actual,
    predicted = knn_base_prediction,
    probability = knn_base_probability
  ),

  evaluate_model(
    model_name = "KNN downsampled",
    actual = evaluation_actual,
    predicted = knn_down_prediction,
    probability = knn_down_probability
  ),

  evaluate_model(
    model_name = "Logistic Regression",
    actual = evaluation_actual,
    predicted = logistic_regression_prediction,
    probability = logistic_regression_probability
  ),

  evaluate_model(
    model_name = "Decision Tree",
    actual = evaluation_actual,
    predicted = decision_tree_prediction,
    probability = decision_tree_probability
  ),

  evaluate_model(
    model_name = "Random Forest",
    actual = evaluation_actual,
    predicted = random_forest_prediction,
    probability = random_forest_probability
  ),

  evaluate_model(
    model_name = "XGBoost",
    actual = evaluation_actual,
    predicted = xgboost_prediction,
    probability = xgboost_probability
  )
) %>%
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 4)
    )
  ) %>%
  arrange(desc(AUC))

kable(model_comparison, caption = "Model comparison")
```

# AUC Comparison For Model Selection

```{r auc-comparison-chart, fig.height=5, fig.width=9}
auc_comparison_plot_data <- model_comparison %>%
  mutate(
    Model = fct_reorder(Model, AUC),
    Selection_Group = if_else(
      Model %in% c(
        "Complement Naive Bayes",
        "Logistic Regression",
        "Random Forest"
      ),
      "Selected / interpreted model",
      "Comparison model"
    )
  )

ggplot(
  auc_comparison_plot_data,
  aes(
    x = Model,
    y = AUC,
    fill = Selection_Group
  )
) +
  geom_col(
    width = 0.72,
    alpha = 0.92
  ) +
  geom_text(
    aes(label = round(AUC, 3)),
    hjust = -0.15,
    size = 3.5,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "Selected / interpreted model" = "#5D3FD3",
      "Comparison model" = "#B57EDC"
    )
  ) +
  scale_y_continuous(
    limits = c(0, 1.05),
    breaks = seq(0, 1, 0.1)
  ) +
  labs(
    title = "AUC Comparison Across Candidate Models",
    subtitle = "Higher AUC indicates stronger ability to separate NEOS from non-NEOS recommendations",
    x = NULL,
    y = "AUC",
    fill = "Model group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(colour = "grey30"),
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

# Confusion Matrices

```{r confusion-matrices}
naive_confusion_matrix <- make_confusion_matrix(
  predicted = naive_prediction,
  actual = evaluation_actual
)

complement_naive_confusion_matrix <- make_confusion_matrix(
  predicted = complement_naive_prediction,
  actual = evaluation_actual
)

knn_base_confusion_matrix <- make_confusion_matrix(
  predicted = knn_base_prediction,
  actual = evaluation_actual
)

knn_down_confusion_matrix <- make_confusion_matrix(
  predicted = knn_down_prediction,
  actual = evaluation_actual
)

logistic_confusion_matrix <- make_confusion_matrix(
  predicted = logistic_regression_prediction,
  actual = evaluation_actual
)

decision_tree_confusion_matrix <- make_confusion_matrix(
  predicted = decision_tree_prediction,
  actual = evaluation_actual
)

random_forest_confusion_matrix <- make_confusion_matrix(
  predicted = random_forest_prediction,
  actual = evaluation_actual
)

xgboost_confusion_matrix <- make_confusion_matrix(
  predicted = xgboost_prediction,
  actual = evaluation_actual
)

confusion_metric_summary <- bind_rows(
  evaluate_model("Naive Bayes", evaluation_actual, naive_prediction, naive_probability),
  evaluate_model("Complement Naive Bayes", evaluation_actual, complement_naive_prediction, complement_naive_probability),
  evaluate_model("KNN baseline", evaluation_actual, knn_base_prediction, knn_base_probability),
  evaluate_model("KNN downsampled", evaluation_actual, knn_down_prediction, knn_down_probability),
  evaluate_model("Logistic Regression", evaluation_actual, logistic_regression_prediction, logistic_regression_probability),
  evaluate_model("Decision Tree", evaluation_actual, decision_tree_prediction, decision_tree_probability),
  evaluate_model("Random Forest", evaluation_actual, random_forest_prediction, random_forest_probability),
  evaluate_model("XGBoost", evaluation_actual, xgboost_prediction, xgboost_probability)
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  arrange(desc(AUC))

kable(
  confusion_metric_summary,
  caption = "Metric check rebuilt from the same confusion matrix inputs"
)

naive_confusion_matrix
complement_naive_confusion_matrix
knn_base_confusion_matrix
knn_down_confusion_matrix
logistic_confusion_matrix
decision_tree_confusion_matrix
random_forest_confusion_matrix
xgboost_confusion_matrix
```

# Model Interpretation And Variable Importance

```{r variable-importance}
# Logistic Regression coefficients
logistic_coefficients <- summary(logistic_regression_model)$coefficients

# Decision Tree variable importance
decision_tree_importance <- varImp(decision_tree_model)

# Random Forest variable importance
random_forest_importance <- varImp(random_forest_model)

# XGBoost variable importance
xgboost_importance <- xgb.importance(
  model = xgboost_model
)

logistic_coefficients
decision_tree_importance
random_forest_importance
xgboost_importance
```

# Save Outputs

```{r save-outputs}
write_csv(model_comparison, file.path(output_dir, "model_comparison.csv"))
write_csv(missing_table, file.path(output_dir, "missing_table.csv"))

saveRDS(naive_model, file.path(output_dir, "naive_bayes_model.rds"))
saveRDS(complement_naive_model, file.path(output_dir, "complement_naive_bayes_model.rds"))
saveRDS(knn_dummy_model, file.path(output_dir, "knn_dummy_model.rds"))
saveRDS(knn_scale_model, file.path(output_dir, "knn_scale_model.rds"))
saveRDS(knn_model_base, file.path(output_dir, "knn_baseline_model.rds"))
saveRDS(knn_model_down, file.path(output_dir, "knn_downsampled_model.rds"))
saveRDS(logistic_regression_model, file.path(output_dir, "logistic_regression_model.rds"))
saveRDS(decision_tree_model, file.path(output_dir, "decision_tree_model.rds"))
saveRDS(xgboost_model, file.path(output_dir, "xgboost_model.rds"))

cat("Outputs saved to:", normalizePath(output_dir), "\n")
```

# Random Forest Max-F1 Threshold Tuning

This section reuses the existing `random_forest_model` and its saved
cross-validation predictions. The model is **not fitted again**. The Max-F1
threshold is selected from out-of-fold training probabilities and is then
compared with the original threshold of 0.50 on the untouched test set.

```{r random-forest-max-f1, message=FALSE, warning=FALSE}
rf_threshold_metrics <- function(actual, probability, threshold) {
  predicted <- probability_to_class(probability, threshold)
  confusion <- confusionMatrix(predicted, actual, positive = "Yes")

  tibble(
    Threshold = threshold,
    Accuracy = unname(confusion$overall["Accuracy"]),
    Sensitivity = unname(confusion$byClass["Sensitivity"]),
    Specificity = unname(confusion$byClass["Specificity"]),
    Precision = unname(confusion$byClass["Pos Pred Value"]),
    F1 = unname(confusion$byClass["F1"]),
    Balanced_Accuracy = unname(
      confusion$byClass["Balanced Accuracy"]
    ),
    Predicted_NEOS_Rate = mean(predicted == "Yes")
  )
}

# Keep only out-of-fold probabilities from the cross-validated mtry selected
# by caret. These probabilities were generated during the existing RF fit.
rf_oof_predictions <- random_forest_model$pred %>%
  filter(mtry == random_forest_model$bestTune$mtry) %>%
  arrange(rowIndex)

rf_threshold_curve <- map_dfr(
  seq(0.01, 0.80, by = 0.01),
  ~ rf_threshold_metrics(
    actual = factor(rf_oof_predictions$obs, levels = c("Yes", "No")),
    probability = rf_oof_predictions$Yes,
    threshold = .x
  )
)

rf_max_f1_result <- rf_threshold_curve %>%
  filter(!is.na(F1)) %>%
  arrange(desc(F1), desc(Sensitivity), desc(Precision)) %>%
  slice_head(n = 1)

rf_max_f1_threshold <- rf_max_f1_result$Threshold[[1]]

kable(
  rf_max_f1_result,
  digits = 4,
  caption = "Random Forest Max-F1 threshold from out-of-fold training predictions"
)
```

## Random Forest Threshold Curve

```{r random-forest-f1-curve, fig.width=9, fig.height=5}
ggplot(rf_threshold_curve, aes(x = Threshold, y = F1)) +
  geom_line(colour = "#5487BE", linewidth = 0.9) +
  geom_vline(
    xintercept = rf_max_f1_threshold,
    colour = "#173B67",
    linetype = "dashed",
    linewidth = 0.8
  ) +
  labs(
    title = "Random Forest Max-F1 Threshold Selection",
    subtitle = paste0(
      "Selected threshold = ",
      round(rf_max_f1_threshold, 3)
    ),
    x = "Probability threshold",
    y = "Out-of-fold F1 score"
  ) +
  theme_minimal()
```

## Test-Set Comparison: 0.50 versus Max-F1

```{r random-forest-threshold-comparison}
rf_evaluation_actual <- factor(
  as.character(test_data$neos_flag),
  levels = c("Yes", "No")
)

rf_roc_result <- roc(
  response = rf_evaluation_actual,
  predictor = random_forest_probability,
  levels = c("No", "Yes"),
  direction = "<",
  quiet = TRUE
)

rf_threshold_comparison <- bind_rows(
  rf_threshold_metrics(
    rf_evaluation_actual,
    random_forest_probability,
    0.50
  ) %>% mutate(Threshold_Method = "Standard 0.50"),
  rf_threshold_metrics(
    rf_evaluation_actual,
    random_forest_probability,
    rf_max_f1_threshold
  ) %>% mutate(Threshold_Method = "Max-F1")
) %>%
  mutate(
    AUC = as.numeric(auc(rf_roc_result)),
    across(where(is.numeric), ~ round(.x, 4))
  ) %>%
  dplyr::select(Threshold_Method, Threshold, AUC, everything())

kable(
  rf_threshold_comparison,
  caption = "Random Forest test performance: threshold 0.50 versus Max-F1"
)

rf_prediction_050 <- probability_to_class(
  random_forest_probability,
  0.50
)

rf_prediction_max_f1 <- probability_to_class(
  random_forest_probability,
  rf_max_f1_threshold
)

rf_confusion_matrix_050 <- confusionMatrix(
  rf_prediction_050,
  rf_evaluation_actual,
  positive = "Yes"
)

rf_confusion_matrix_max_f1 <- confusionMatrix(
  rf_prediction_max_f1,
  rf_evaluation_actual,
  positive = "Yes"
)

rf_confusion_matrix_050
rf_confusion_matrix_max_f1
```

## Sensitivity and Precision Trade-off

```{r random-forest-threshold-chart, fig.width=9, fig.height=5}
rf_threshold_comparison %>%
  dplyr::select(Threshold_Method, Sensitivity, Precision) %>%
  pivot_longer(
    cols = c(Sensitivity, Precision),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  ggplot(aes(x = Metric, y = Value, fill = Threshold_Method)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.68) +
  geom_text(
    aes(label = scales::percent(Value, accuracy = 0.1)),
    position = position_dodge(width = 0.75),
    vjust = -0.35,
    size = 3.5
  ) +
  scale_y_continuous(
    labels = scales::percent_format(),
    limits = c(0, 1.05)
  ) +
  scale_fill_manual(values = c(
    "Standard 0.50" = "#9BBAD5",
    "Max-F1" = "#173B67"
  )) +
  labs(
    title = "Random Forest Operating Threshold Trade-off",
    x = NULL,
    y = "Test-set performance",
    fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

# Random Forest SHAP Explanation

SHAP explains how each feature contributes to the existing Random Forest's
predicted NEOS probability. It explains model behaviour and does not establish
causality. To control runtime, SHAP uses 100 test observations and 100 training
background observations.

```{r random-forest-shap, message=FALSE, warning=FALSE}
set.seed(SEED)

rf_train_features <- train_data %>% dplyr::select(-neos_flag)
rf_test_features <- test_data %>% dplyr::select(-neos_flag)

rf_shap_explanation_data <- rf_test_features %>%
  slice_sample(n = min(100, n()))

rf_shap_background_data <- rf_train_features %>%
  slice_sample(n = min(100, n()))

rf_neos_probability <- function(object, newdata) {
  predict(object, newdata = newdata, type = "prob")[, "Yes"]
}

rf_shap_result <- kernelshap(
  object = random_forest_model,
  X = rf_shap_explanation_data,
  bg_X = rf_shap_background_data,
  pred_fun = rf_neos_probability,
  exact = FALSE,
  hybrid_degree = 1,
  tol = 0.02,
  max_iter = 20,
  verbose = FALSE,
  seed = SEED
)

rf_shap_values <- as.data.frame(rf_shap_result$S)
```

## Global SHAP Importance

```{r random-forest-global-shap, fig.width=9, fig.height=6}
rf_global_shap_importance <- tibble(
  Feature = names(rf_shap_values),
  Mean_Absolute_SHAP = map_dbl(
    rf_shap_values,
    ~ mean(abs(.x), na.rm = TRUE)
  )
) %>%
  arrange(desc(Mean_Absolute_SHAP))

kable(
  rf_global_shap_importance %>% slice_head(n = 15),
  digits = 4,
  caption = "Top Random Forest drivers by mean absolute SHAP value"
)

rf_global_shap_importance %>%
  slice_head(n = 15) %>%
  ggplot(aes(
    x = fct_reorder(Feature, Mean_Absolute_SHAP),
    y = Mean_Absolute_SHAP
  )) +
  geom_col(fill = "#173B67", width = 0.68) +
  coord_flip() +
  labs(
    title = "Random Forest Global SHAP Importance",
    subtitle = "Larger values indicate greater influence on predicted NEOS probability",
    x = NULL,
    y = "Mean absolute SHAP value"
  ) +
  theme_minimal()
```

## Individual SHAP Explanation

```{r random-forest-individual-shap}
rf_individual_shap <- tibble(
  Feature = names(rf_shap_values),
  SHAP_Value = as.numeric(rf_shap_values[1, ])
) %>%
  mutate(Direction = if_else(
    SHAP_Value >= 0,
    "Pushes toward NEOS",
    "Pushes away from NEOS"
  )) %>%
  arrange(desc(abs(SHAP_Value))) %>%
  slice_head(n = 10)

kable(
  rf_individual_shap,
  digits = 4,
  caption = "Top SHAP contributions for one example test observation"
)
```

# Save Random Forest Threshold and SHAP Outputs

```{r save-random-forest-threshold-shap}
write_csv(
  rf_threshold_comparison,
  file.path(output_dir, "random_forest_threshold_050_vs_max_f1.csv")
)

write_csv(
  rf_threshold_curve,
  file.path(output_dir, "random_forest_f1_threshold_curve.csv")
)

write_csv(
  rf_global_shap_importance,
  file.path(output_dir, "random_forest_global_shap_importance.csv")
)

saveRDS(
  rf_shap_result,
  file.path(output_dir, "random_forest_shap_result.rds")
)

cat("Random Forest threshold and SHAP outputs saved to:", normalizePath(output_dir), "\n")
```
